### 2018-06-03 doc/manuscript.Rmd
added citations, minor edits, and fixes for finished draft to send to tony

In the present study, we found that in a Canadian cohort at risk for T2D,
several specific TGFA and groups of TGFA were strongly associated
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, the TGFA myristic acid (14:0), 7-tetradecenoic acid 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TGFA were not
associated with beta-cell function, two fatty acids, palmitic acid (16:0) and
*cis*-vaccenic acid (18:1n-7), were associated negatively and positively,
respectively, with measures of beta-cell function. Using PLS, we also found that four
TGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
strongly predicted lower insulin sensitivity.

@@

There are a few possible explanations to these findings. Fatty acids, and thus
the TGFA composition, derive from two sources: dietary fat and de novo
lipogenesis (DNL). Adipose tissue also supplies circulating fatty acids,
however, these also ultimately originate from the two sources. Specific dietary
fat and carbohydrates can influence DNL activity {{cite}}. Determining the
specific source of fatty acids is extremely difficult to ascertain outside of
highly controlled experimental settings. However, there are sign that suggest
one source over another.

The link DNL activity and levels of specific fatty acids has been examined in
several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. {{talk about Yki-.. study}}.  In our findings, the four
fatty acids were highly correlated amongst each other, in addition to clustering
together on their negative association with insulin sensitivity. If these fatty
acids derived from mainly the diet, we would anticipate seeing more correlation
with the 18:0, 18:1, and 18:2 fatty acids, as these fatty acids often share a
dietary source (e.g. meat) {{cite?}} with the four clustering fatty acids. As our
results do not show this, this suggests the common source being DNL and that the
fatty acids may have been produced from higher DNL activity since, again, there
was no strong correlation with 18:0 and 18:1 fatty acids, which would be the
next stage DNL via elongation and/or desaturation of 16:0 and 16:1 fatty acids.

A higher DNL activity may result from a higher carbohydrate intake as a means to
control blood glucose [@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a;
@Parks1999a; @Chehade2013a; @Verges2015a], which could increase hepatic fat
stores and increase the amount of TG in circulation. In fact, we found that
higher proportions of these four TGFA also highly correlated with a higher
concentration of clinical TG. Several studies have shown a link between higher
estimated DNL and an increased risk for metabolic dysfunction [@Ma2015a;
@Zong2013a; @Lankinen2015a; @Kroger2011a]; how DNL may influence metabolic
dysfunction is not well understood. Possible explanations may be that higher DNL
produces more of certain fatty acids or that higher DNL increases circulating TG,
which itself is well documented to contribute to metabolic dysfunction {{cite?}}.

Regardless of the exact source of these fatty acids, our results in addition to
the available scientific evidence emphasizes the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions {{cite}}. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly predicted lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear
from previous cross-sectional studies due to the physiological feedback
mechanisms involved. For example, while greater DNL and higher TGFA may promote
muscle insulin resistance, the reverse may also be true [@Flannery2012a]. Higher
insulin resistance may encourage TGFA to remain in circulation longer {{cite?}}.
As we found no interaction by time of TGFA on insulin sensitivity, this study
cannot determine the exact role of the feedback mechanism. However, combining the
with the lack of time interaction and the consistent negative association
without the time interaction, these results at least suggest that the feedback
mechanism may not be strongly influential and that TGFA may predict insulin
sensitivity over a six year period. Given the complex biological mechanisms and
feedback loops involved, disentangling whether insulin sensitivity influences
TGFA more strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

@@

Our study has potential limitations that need to be considered when interpreting
the results. Firstly, this is an observational cohort and as such there may be
some residual confounding we were not able to control for or were unaware of. 
However, we have taken extensive, empirically based precautions in identifying
potential confounders and mediators through the use of the DAG modeling, relying
on previous literature, and through information criteria model fit comparison methods.

Only fasting TGFA were quantified, and only at the baseline visit. TGFA 
composition can fluctuate substantially throughout the day, so in order to 
control for this, PROMISE participants came for the clinic visit in the morning
and fasted. There is some evidence to suggest that fasting TG is better able
to discriminate diabetes cases compared to a post-prandial state [@Rhee2011a].
Because TGFA were only measured at the baseline visit, we cannot
investigate whether there are concomitant changes in TGFA and the metabolic
measures over time. However, to optimally use GEE to analyze the data and for
interpretation, we used the model to infer that a given value of TGFA could
predict values of insulin sensitivity or beta-cell function over a 6 year
period. This in our view is a strength of our analysis, as it reduces the chance
of reverse causality given the tight integration of the glucose and fatty acid
metabolism pathways, as well as maximizes the specific usage of the GEE
modeling.

>>>>

In the present study, we found that in a Canadian cohort at risk for T2D,
several specific TGFA and groups of TGFA were strongly associated
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, the TGFA myristic acid (14:0), 7-tetradecenoic acid 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively associated with lower insulin sensitivity. While most TGFA were not
associated with beta-cell function, three fatty acids, palmitic acid (16:0),
*cis*-vaccenic acid (18:1n-7), and eicosadienoic acid (20:2n-6) were associated negatively and positively,
respectively, with measures of beta-cell function. Using PLS, we also found that four
TGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
strongly predicted lower insulin sensitivity.

@@

There are a few possible explanations to these findings. Circulating 
TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
of TGFA is extremely difficult to ascertain outside of highly controlled
experimental settings. Many previous studies that have examined DNL
have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
study found that palmitoleic acid (16:1n-7), directly measured DNL using
isotopes, and liver fat were all highly correlated with each other (r>0.50),
suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
end of the study, the carbohydrate group had higher DNL as well as an increase
in liver fat, though the saturated fat group had the highest increase in liver
fat.

This link between carbohydrate intake and DNL activity has been well documented
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].
In our findings, the four fatty acids were highly positively correlated
amongst each other and negatively or neutrally with all other TGFA, in addition to
clustering together on their negative association with insulin sensitivity. This
may suggest that greater DNL activity may be the source of these TGFA. Several
studies have shown a link between higher estimated DNL and an increased risk for
metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; how
DNL may influence metabolic dysfunction is not well understood. Possible
reasons may be that higher DNL produces more of certain fatty acids or that
higher DNL increases circulating TG, which itself is well documented to
contribute to metabolic dysfunction and which we found in our study of a high
positive correlation between the four TGFA and clinical TG.

Regardless of the exact source of these fatty acids, our results in addition to
the available scientific evidence emphasizes the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly associated with lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear
from previous cross-sectional studies due to the physiological feedback
mechanisms involved. For example, while higher TGFA may promote
muscle insulin resistance, the reverse may also be true [@Flannery2012a].
As we found no interaction by time of TGFA on insulin sensitivity, this study
cannot determine the exact role of the feedback mechanism. However, combining the
with the lack of time interaction and the consistent negative association
without the time interaction, these results at least suggest that the feedback
mechanism may not be strongly influential and that TGFA may predict insulin
sensitivity at least over a six year period. Given the complex biological mechanisms and
feedback loops involved, disentangling whether insulin sensitivity influences
TGFA more strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

@@

Our study has potential limitations that need to be considered when interpreting
the results. Firstly, this is an observational cohort and as such there may be
some residual confounding we were not able to control for or were unaware of. 
However, we have taken extensive, empirically based precautions in identifying
potential confounders and mediators through the use of the DAG modeling, relying
on previous literature, and through information criteria model fit comparison methods.
Only fasting TGFA were quantified, and only at the baseline visit. TGFA 
composition can fluctuate substantially throughout the day, so in order to 
control for this, PROMISE participants came for the clinic visit in the morning
and fasted. There is some evidence to suggest that fasting TG is better able
to discriminate diabetes cases compared to a post-prandial state [@Rhee2011a].
Because TGFA were only measured at the baseline visit, we cannot
investigate whether there are concomitant changes in TGFA and the metabolic
measures over time. However, to optimally use GEE to analyze the data and for
interpretation, we used the model to infer that a given value of TGFA could
predict values of insulin sensitivity or beta-cell function over a 6 year
period. This in our view is a strength of our analysis, as it reduces the chance
of reverse causality given the tight integration of the glucose and fatty acid
metabolism pathways, as well as maximizes the specific usage of the GEE
modeling.
